---
title: "_daxx_ and innate immunity in _Drosophila_"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    highlight: haddock
---

## Computing tools

### Libraries
```{r, message = FALSE, warning = FALSE}
# data handling
library(readxl)
library(tidyverse)
library(dplyr)
# stats
library(survival)
library(survminer)
# for plotting
library(ggthemes) # italics in the figures
library(ggtext) # italics in the figures
library(RColorBrewer)
#library(viridis)
```

### Custom function `analyse_spreadsheet`
```{r, message = FALSE, warning = FALSE}
devtools::source_url('https://raw.githubusercontent.com/jdenavascues/Survival_functions/main/infection.R')
```

## Experiment parameters

* Start date: DD-MM-YYYY
* Lab members: `z`
* Genotypes:
    * e.g. immunodeficient ( _key^1^_)
    * _x-Gal4, UAS-y^RNAi-GD....^_
    * _x-Gal4, UAS-ctrl^RNAi^_
* Pathogen: e.g. _S. marcescens_ (Db11?, OD~600~? volume?)
* Infection method: (oral, prickling, injection)

## Overall log-rank stats and plotting

### Load and organise data
```{r,  echo = FALSE, message = FALSE, warning = FALSE}
filepath <- file.path('/Users/JQ/Documents/__REPOS/GitHub/Survival_functions',
                      'Infection_data_example.xlsx')
data <- analyse_spreadsheet(filepath, cph=TRUE)
```



Prepare the graph:

```{r,  echo = FALSE, message = FALSE, warning = FALSE}
palette <- c('#85C1E9', '#999999', '#E74C3C', #'#F1C40F'
             '#4444DD', '#000000', '#A2103C') # '#F39C12'
line_type <- c(rep("solid",3),rep("twodash",3))
title <- 'Susceptibility to *S. marcescens*'
g <- ggsurvplot(surv_fit(Surv(hour, event) ~ genotype + treatment,
                         data = data),
                pval = TRUE,                       # adds pval
                conf.int = TRUE,                   # adds 95% confidence interval
                conf.int.alpha = 0.1,              # makes it more transparent
                surv.median.line = "v",           # adds median points
                palette = palette,
                xlab = "Time (hours)",              # remove x axis label
                ylab = "Survival Probability",     # y axis label
                linetype = line_type,
                title = title,
                font.legend = c(7),               # legend size
                legend = "bottom")                  # legend position
```

#### To prepare the legend:

Let's have a look at our strata (by default organised as `genotype:treatment` in alphabetical order): 
```{r,  echo = FALSE, message = FALSE, warning = FALSE}
levels(g$data.survtable$strata)
```
But I'd like to present the strata as here:
```{r,  echo = FALSE, message = FALSE, warning = FALSE}
exgrid <- expand.grid(unique(sort(data$genotype, decreasing = TRUE)),
                      unique(sort(data$treatment, decreasing = TRUE)))
exgrid <- exgrid[order(exgrid$Var2),]
lab_strata <- paste(exgrid$Var1, exgrid$Var2, sep=" | ")
lab_strata
```
```{r}
lab_strata <- c("*myoIA>ctrl^KD^* | Sucrose",
                "*myoIA>Etl1^KD^* | Sucrose",
                "*ken^1^* | Sucrose",
                "*myoIA>ctrl^KD^* | S.marcescens",
                "*myoIA>Etl1^KD^* | S.marcescens",
                "*ken^1^* | S.marcescens")
lab_strata
```
to match the colours and the line types like this:

```{r,  echo = FALSE, message = FALSE, warning = FALSE, fig.height = 1, fig.width=2.5}
x <- rep(c(1:20), 6)
variable <- unlist(lapply(lab_strata, rep, 20))
value <- c( 5+rnorm(20)/5, 4+rnorm(20)/5, 3+rnorm(20)/5,
            2+rnorm(20)/5, 1+rnorm(20)/5, rnorm(20)/5 )
df <- data.frame(x, variable, value)
d <- ggplot(df, aes(x=x, y=value, group=variable, colour=variable )) +
  geom_line(aes(linetype=variable, colour=variable)) +
  scale_linetype_manual(breaks=lab_strata, values=line_type, labels=LETTERS[1:6]) +
  scale_colour_manual(breaks=lab_strata, values=palette, labels=LETTERS[1:6]) +
  theme(legend.text=element_text(size=5),
        legend.title = element_text(size=6),
        legend.key.height = unit(1/6, 'in'))
d
```

To match what we want, the order of the legends, colours and line types should be like:
```{r,  echo = FALSE, message = FALSE, warning = FALSE}
lab_strata[c(1,3,2,4,6,5)]
```
And the reverse order would be:
```{r,  echo = FALSE, message = FALSE, warning = FALSE}
l <- levels(g$data.survtable$strata)[c(6,2,4,5,1,3)]
l
```
Now we can start with the preliminary plot:

```{r,  echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3, fig.width=5}
g$plot +
  scale_colour_manual(name='', breaks=l, values=palette,
                      labels=lab_strata[c(1,3,2,4,6,5)],
                      guide=guide_legend(nrow = 3)) +
  scale_fill_manual(name='', breaks=l, values=palette,
                    labels=lab_strata[c(1,3,2,4,6,5)],
                    guide=guide_legend(nrow = 3)) +
  scale_linetype_manual(name='', breaks=l, values=line_type,
                        labels=lab_strata[c(1,3,2,4,6,5)],
                        guide=guide_legend(nrow = 3)) +
  theme(legend.text=element_markdown(),
        plot.title = element_markdown())
```

It seems that the only susceptible flies are the _ken^1^_