---
title: "Evaluation of pupation time with _drug_ and the _mut^1^_ allele"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    highlight: haddock
---

## 1. Computing tools

### Libraries
```{r, message = FALSE, warning = FALSE}
# data handling
library(readxl)
# stats
library(survival)
library(survminer)
# for plotting
library(ggthemes) # italics in the figures
library(ggtext) # italics in the figures
library(RColorBrewer)
```

### Custom function `analyse_spreadsheet`
```{r, message = FALSE, warning = FALSE}
devtools::source_url('https://raw.githubusercontent.com/jdenavascues/Survival_functions/main/development.R')
#devtools::source_url('https://raw.githubusercontent.com/jdenavascues/Survival_functions/master/development.R')
```

## 2. Experiment parameters

* Start date: 21.11.2022
* Lab members: Keri-Ann
* Genotypes: _wild-type_, _mut^1^_
* Treatment: drug 0mM / 50mM
* Treatment method: mixed in melted food, with E133 dye.

## 3. Statistical analysis

### 3.1.Prepare your data

First load your data. You should have recorded it in a copy of the file
`Pupariation_data_template.xlsx`, in this form:

|   Date    |  Time  | Event | Treatment | Dose | Dose_unit | Genotype    | Replicate |
|:---------:|:------:|------:|:----------|-----:|:----------|:------------|----------:|
| DD.MM.YYYY | hh:mm |   0   |   *drug*  |   0  |   *mM*    | *wild-type* |    1      |
| DD.MM.YYYY | hh:mm |   0   |   *drug*  |   0  |   *mM*    | *wild-type* |    2      |
| DD.MM.YYYY | hh:mm |   0   |   *drug*  |   0  |   *mM*    | *wild-type* |    3      |
| DD.MM.YYYY | hh:mm |   1   |   *drug*  |  50  |   *mM*    | *wild-type* |    1      |
| DD.MM.YYYY | hh:mm |   0   |   *drug*  |  50  |   *mM*    | *wild-type* |    2      |
| DD.MM.YYYY | hh:mm |   0   |   *drug*  |  50  |   *mM*    | *wild-type* |    3      |
etc.

This should be in the 'data' spreadsheet, with a 'metadata' spreadsheet specifying the characteristics of the experiment.
This is a convenient way for humans to capture the data (time-based), while the `survival` statistics need an event-based recording.
This is done with the function `load_devtime_data`:
```{r, message = FALSE, warning = FALSE}
# YOU WILL NEED TO ADAPT THESE LINES FOR R TO WORK WITH YOUR FILE IN YOUR COMPUTER
filepath <- file.path('/Users/JQ/Documents/_CODE REPOS/GitHub/Survival_functions',
                      'Pupariation_data_example.xlsx')
data <- load_devtime_data(filepath)
head(data, 5)
```

### 3.2.Data plotting and analysis

#### 3.2.1-Analysis of replicates/outliers

These experiments are sometimes temperamental, and have large batch variations, so first let us make sure that the replicates show homogeneous behaviour.
Let's start with _wild-type_ flies:
```{r, message = FALSE, warning = FALSE}
colrs <- brewer.pal(12,"Paired")[c(2,8)]
palette <- c(rep(colrs[1],4), rep(colrs[2],4))
lty_all <- c("solid", "twodash", "dotted", "longdash", "dashed", "dotdash")
line_type <- rep(lty_all[c(1,2,3,6)], 3)
title <- 'Replicates of *wild-type* flies'
legend_labs <- c("Control, R1", "Control, R2", "Control, R3", "Control, R4",
                 "drug 50mM, R1", "drug 50mM, R2", "drug 50mM, R3", "drug 50mM, R4")
model <- survfit(Surv(hour, event) ~ dose + replicate,
                 data=subset(data, genotype=='*wild-type*'))

g <- ggsurvplot(model, data=subset(data, genotype=='*wild-type*'),
                pval = TRUE,                             # adds pval
                fun = "event",                           # invert
                conf.int = TRUE,                         # adds 95% confidence interval
                conf.int.alpha = 0.1,                    # makes it more transparent
                surv.median.line = "hv",                 # adds median points
                palette = palette,
                linetype = line_type,
                title = title,
                xlab = "Time (Hours)",                   # remove x axis label
                ylab = "Pupariation Probability",        # y axis label
                legend.title = "Replicates",             # legend title
                font.legend = c(7),                      # legend size
                legend = "bottom")                       # legend position

l <- levels(g$data.survtable$strata)

g$plot +
  scale_linetype_manual(breaks=l, values=line_type, labels=legend_labs, guide=guide_legend(nrow = 2)) +
  scale_colour_manual(breaks=l, values=palette, labels=legend_labs, guide=guide_legend(nrow = 2)) +
  scale_fill_manual(breaks=l, values=palette, labels=legend_labs, guide=guide_legend(nrow = 2)) +
  theme(legend.text=element_markdown(),
        plot.title = element_markdown())
```

It does look like drug-50mM replicate 2 should be eliminated from the analysis.

Now with _mut^1^_ flies:
```{r}
colrs <- brewer.pal(12,"Paired")[c(2,8)]
palette <- c(rep(colrs[1],3), rep(colrs[2],3))
lty_all <- c("solid", "twodash", "dotted", "longdash", "dashed", "dotdash")
line_type <- rep(lty_all[c(1,2,3)], 2)
title <- 'Replicates of *mut^1^* flies'
legend_labs <- c("Control, R1", "Control, R2", "Control, R3",
                 "drug 50mM, R1", "drug 50mM, R2", "drug 50mM, R3")
model <- survfit(Surv(hour, event) ~ dose + replicate,
                 data=subset(data, genotype=='*mut^1^*'))

g <- ggsurvplot(model, data=subset(data, genotype=='*mut^1^*'),
                pval = TRUE,                             # adds pval
                fun = "event",                           # invert
                conf.int = TRUE,                         # adds 95% confidence interval
                conf.int.alpha = 0.1,                    # makes it more transparent
                surv.median.line = "hv",                 # adds median points
                palette = palette,
                linetype = line_type,
                title = title,
                xlab = "Time (Hours)",                   # remove x axis label
                ylab = "Pupariation Probability",        # y axis label
                legend.title = "Replicates",             # legend title
                font.legend = c(7),                      # legend size
                legend = "bottom")                       # legend position

l <- levels(g$data.survtable$strata)

g$plot +
  scale_linetype_manual(breaks=l, values=line_type, labels=legend_labs, guide=guide_legend(nrow = 2)) +
  scale_colour_manual(breaks=l, values=palette, labels=legend_labs, guide=guide_legend(nrow = 2)) +
  scale_fill_manual(breaks=l, values=palette, labels=legend_labs, guide=guide_legend(nrow = 2)) +
  theme(legend.text=element_markdown(),
        plot.title = element_markdown())
```
This is ok.
So we can go ahead with everything but replicate 2 of wild-type, drug-50mM.

#### 3.2.2-Full model

```{r, message = FALSE, warning = FALSE}
palette <- brewer.pal(12,"Paired")[c(1,2,7,8)]
lty_all <- c("solid", "twodash", "dotted", "longdash", "dashed", "dotdash")
line_type <- lty_all[c(1,4,1,4)]
title <- 'Effect of *mut^1^* on response to 50mM drug'
legend_labs <- c("*wild-type*, control", "*wild-type*, 50mM drug",
                 "*mut^1^*, control", "*mut^1^*, 50mM drug")
model <- survfit(Surv(hour, event) ~ genotype + dose,
                 data=subset(data, !(genotype=='*wild-type*' & dose==50 & replicate==2)))

g <- ggsurvplot(model,
                data=subset(data, !(genotype=='*wild-type*' & dose==50 & replicate==2)),
                pval = TRUE,                             # adds pval
                fun = "event",                           # invert
                conf.int = TRUE,                         # adds 95% confidence interval
                conf.int.alpha = 0.1,                    # makes it more transparent
                surv.median.line = "hv",                 # adds median points
                palette = palette,
                linetype = line_type,
                title = title,
                xlab = "Time (Hours)",                   # remove x axis label
                ylab = "Pupariation Probability",        # y axis label
                legend.title = "Treatment",             # legend title
                font.legend = c(10),                      # legend size
                legend = "bottom")                       # legend position

l <- levels(g$data.survtable$strata)

g$plot +
  scale_linetype_manual(breaks=l, values=line_type, labels=legend_labs, guide=guide_legend(nrow = 2)) +
  scale_colour_manual(breaks=l, values=palette, labels=legend_labs, guide=guide_legend(nrow = 2)) +
  scale_fill_manual(breaks=l, values=palette, labels=legend_labs, guide=guide_legend(nrow = 2)) +
  theme(legend.text=element_markdown(),
        plot.title = element_markdown())
```
```{r, echo=FALSE}
ggsave('mut_drug.pdf', plot=last_plot(),
       width = 5,
       height = 4,
       units = "in",
       dpi = 300)
```

```{r}
## get median values with confidence interval
surv_median(model)
```

```{r, warning= = FALSE, message = FALSE}
ratedata <- load_devtime_data(filepath, censor=TRUE)
tail(ratedata)
```

```{r, warning= = FALSE, message = FALSE}
pupas <- ratedata %>%
  subset(!(replicate==2 & genotype=='*wild-type*' & dose==50)) %>%
  filter(event==1) %>%
  group_by(genotype, dose, replicate) %>%
  summarise(n())
rep_size <- 50
pupas$percent <- pupas$`n()`*100/rep_size
pupas <- pupas[c('genotype', 'dose', 'percent')]
pupas
```
```{r, message = FALSE, warning = FALSE}
meanlines <- summarise(pupas, mean=mean(percent),
                       by=interaction('genotype', 'dose'), .groups='keep')
meanlines$x <- interaction(meanlines$genotype, meanlines$dose, sep=" | ")

g <- ggplot(pupas, aes(x = interaction(genotype, dose, sep=" | "), y = percent)) +
  geom_dotplot(binaxis='y', stackdir='center', stackratio=1.5, dotsize=1.2) +
  geom_segment(data=meanlines,
               aes(x=as.numeric(x)-0.5, xend=as.numeric(x)+0.5,
                   y=mean, yend=mean),
               colour='red',
               size=1) +
  labs(x = "Genotype | drug (mM)", y="Pupation rate") +
  theme(axis.text.x=element_markdown(),
        axis.text.y=element_markdown())
g
```

```{r, echo=FALSE}
ggsave('pupation_rates.pdf', plot=last_plot(),
       width = 5,
       height = 4,
       units = "in",
       dpi = 300)
```