---
title: "DAXX, SCFAs and innate immunity in _Drosophila_"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    highlight: haddock
---

## Computing tools

### Libraries
```{r, message = FALSE, warning = FALSE}
# data handling
library(readxl)
library(tidyverse)
library(magrittr)
# stats
library(survival)
library(survminer)
# for aesthetics/plotting
library(ggthemes) # italics in the figures
library(ggtext) # italics in the figures
require(knitr)
require(kableExtra)
```

### Custom function `analyse_spreadsheet`
```{r, message = FALSE, warning = FALSE}
devtools::source_url('https://raw.githubusercontent.com/jdenavascues/Survival_functions/main/infection.R')
```

# Experiment parameters

* Start date: 04/03/2024
* Lab members: `Oleksander, Catherine and Joanne` (Frontrunners 2023/24)
* Genotypes:
    * _daxx^NP4778^_
    * _Dahomey wild types_
* Pathogen: e.g. _S. marcescens_ Db11 (OD~600~=100)
* Infection method: oral

# Preparing the data

## Loading the data

Quick look at the data:
```{r,  echo = FALSE, message = FALSE, warning = FALSE}
filepath <- file.path('/Users/JQ/Documents/__RESEARCH/__PROJECTS/_SMARCAD+CrA/Infection/_Cath_Joanne_Olek',
                      'survival data collection template .xlsx')
data <- read_xlsx(filepath, skip = 1)
# Remove unnecessary rows and columns
data <- data[-c(1,2), ]
head(data)
```

We need to read the timepoints, putting together the date from one row with the time from the other.
Thenan rename columns:
```{r}
# read time point info from excel file
data2 <- read_xlsx(filepath,
                   range='G3:V4',
                   col_names=paste0('t',as.character(1:16)),
                   col_types=rep('date',16)) %>%
  tibble::rownames_to_column() %>%  
  pivot_longer(-rowname) %>% 
  pivot_wider(names_from=rowname, values_from=value) %>%
  mutate(datetime = str_c(as.Date(`1`), strftime(`2`, format="%H:%M:%S"), sep=' '))
# rename columns to make them more descriptive
colnames(data) <- c("Genotype", "Treatment_1", "Sex", "Treatment_2", "Replicate", data2$datetime)
data %<>% mutate(across(6:length(colnames(data)), ~ as.integer(.x)))
head(data)
```

## A bit of data wrangling

Now we need this in long form:
```{r}
data %<>%
  pivot_longer(cols = 6:21, names_to = "datetime", values_to = "deaths") %>%
  mutate(date = as.Date(datetime)) %>%
  mutate(time = strftime(datetime, format="%H:%M:%S")) %>%
  select(!datetime)
head(data)
```

In principle, this should be enough to start applying the `analyse_spreadsheet` function from my repository.
However, this has two limitations:

* it is designed for ONE treatment, not two at the same time
* it is designed for the same `rep_size` for all samples and replicates

Quick look at the replicate sizes:
```{r}
rep_size <- read_xlsx(filepath, sheet='rep_size')
colnames(rep_size) <- c('genotype', 'treatment_1', 'sex', 'treatment_2', 'replicate', 'size')
head(rep_size)
```

Now we can provide the wrapper function with all we need for a quick prelim work
```{r}
data <- analyse_spreadsheet(data, rep_size=rep_size, cph=TRUE)
```

Now we have a data.frame with all the original data but also:
- all data checked for integrity and internal consistency
- 'repaired' data from lost corpses eaten up by larvae
- additional censoring of all surviving flies
- format compatible with `Survival` package
- and a quick preliminary view that Cox proportional hazards are not respected
```{r}
tb <- knitr::kable(data, format = "html")
kableExtra::kable_styling(tb, "striped", position = "left", font_size = 10,
              html_font = '"Arial", arial, helvetica, sans-serif')
```

# Hypothesis testing

First let us explore different hypotheses:
- Are _Pseudomonas entomophila_ really killing flies?
- If _P.ent._ are pathogenic,
  - are _Drosophila_ males and females different in their susceptibility?
  - are _daxx^NP4778^_ mutants more susceptible than _Dahomey_ wild-types?
  - does raising the larvae in crotonic acid affect the adult immuno-competence?
- Do any of the previous variables interact, i.e. susceptibility only changes in certain combinations of the variables?

We can do this using `survival` and `survminer` packages:
```{r}
# from `survival`:
pathogenicity <- surv_fit( Surv(time_to_event, event_type) ~ treatment_2, data = data) # do Pseudomonas kill flies?
# from `survminer`:
surv_pvalue(pathogenicity, method = "S1",
            test.for.trend = FALSE, combine = FALSE)[1:3]
```
So, bacteria do kill flies. Not with spectacular confidence, but they do. (Maybe they will if we focus on specific groups).

So, does sex impact survival?
```{r}
sexspec <- surv_fit( Surv(time_to_event, event_type) ~ sex,
                     data = data) # does sex affect survival?
# from `survminer`:
surv_pvalue(sexspec, method = "1",
            test.for.trend = FALSE, combine = FALSE)[1:3]
ggsurvplot(sexspec)
```
It does. But, does it affect immunocompetence, or just `survival`? It seems to be a much stronger variable than infection itself! We cannot test formally for variable interactions, but we can try to see "qualitative" interactions: do effects go away if we consider only a subset of the data? So, does sex matter for the infected flies?
```{r}
sexspec_inf <- surv_fit( Surv(time_to_event, event_type) ~ sex,
                     data = filter(data, treatment_2=='P.ent.')) # does sex affect survival?
surv_pvalue(sexspec_inf, method = "1",
            test.for.trend = FALSE, combine = FALSE)[1:3]
ggsurvplot(sexspec_inf)
```
Ok, it does - but is it the same for non-infected flies?
```{r}
sexspec_mock <- surv_fit( Surv(time_to_event, event_type) ~ sex,
                     data = filter(data, treatment_2=='vehicle')) # does sex affect survival?
surv_pvalue(sexspec_mock, method = "1",
            test.for.trend = FALSE, combine = FALSE)[1:3]
ggsurvplot(sexspec_mock)
```
Even more so!!! Maybe this has to do with the genotype?
Then we should probably explore this more systematically:
```{r}
# also from survminer
res <- pairwise_survdiff(Surv(time_to_event, event_type) ~ 
                           sex + genotype + treatment_2,
                         data = data)
# capture as dataframe
dt <- as.data.frame(round(res$p.value,3))
names(dt) <- str_remove_all(names(dt), "sex=|genotype=|treatment_2=")
rownames(dt) <- str_remove_all(rownames(dt), "sex=|genotype=|treatment_2=")
dt %>%
#  mutate(across(everything(), ~replace_na(., 1))) %>%
  mutate(across(everything(),
                ~ case_when(. >= 0.05 ~ cell_spec(., color = "#777777"),
                            (. < 0.05 & 0.01 <= .) ~ cell_spec(., background = "#ffaacc"),
                            (. < 0.01 & 0.001 <= .) ~ cell_spec(., background = "#ee7799"),
                            . < 0.001 ~ cell_spec(., background = "#ee5577")))) %>%
  kbl(booktabs = TRUE,
      linesep = "",
      escape = FALSE) %>%
  kable_minimal(full_width = FALSE)
```


# Plotting

```{r,  echo = FALSE, message = FALSE, warning = FALSE}
# palette <- c('#85C1E9', '#999999', '#E74C3C', #'#F1C40F'
#              '#4444DD', '#000000', '#A2103C') # '#F39C12'
#line_type <- c(rep("solid",3),rep("twodash",3))
title <- 'Susceptibility to *S. marcescens*'
model <- surv_fit(Surv(time_to_event, event_type) ~ genotype + treatment_2,
                  data = filter(data, sex=='F'))
g <- ggsurvplot(model,
                pval = TRUE,                       # adds pval
                conf.int = TRUE,                   # adds 95% confidence interval
                conf.int.alpha = 0.1,              # makes it more transparent
                surv.median.line = "v",           # adds median points
                #palette = palette,
                xlab = "Time (hours)",              # remove x axis label
                ylab = "Survival Probability",     # y axis label
                #linetype = line_type,
                title = title,
                font.legend = c(7),               # legend size
                legend = "bottom")                  # legend position
g
```

#### To prepare the legend:

Let's have a look at our strata (by default organised as `genotype:treatment` in alphabetical order): 
```{r,  echo = FALSE, message = FALSE, warning = FALSE}
levels(g$data.survtable$strata)
```
But I'd like to present the strata as here:
```{r,  echo = FALSE, message = FALSE, warning = FALSE}
exgrid <- expand.grid(unique(sort(data$genotype, decreasing = TRUE)),
                      unique(sort(data$treatment, decreasing = TRUE)))
exgrid <- exgrid[order(exgrid$Var2),]
lab_strata <- paste(exgrid$Var1, exgrid$Var2, sep=" | ")
lab_strata
```
```{r}
lab_strata <- c("*myoIA>ctrl^KD^* | Sucrose",
                "*myoIA>Etl1^KD^* | Sucrose",
                "*ken^1^* | Sucrose",
                "*myoIA>ctrl^KD^* | S.marcescens",
                "*myoIA>Etl1^KD^* | S.marcescens",
                "*ken^1^* | S.marcescens")
lab_strata
```
to match the colours and the line types like this:

```{r,  echo = FALSE, message = FALSE, warning = FALSE, fig.height = 1, fig.width=2.5}
x <- rep(c(1:20), 6)
variable <- unlist(lapply(lab_strata, rep, 20))
value <- c( 5+rnorm(20)/5, 4+rnorm(20)/5, 3+rnorm(20)/5,
            2+rnorm(20)/5, 1+rnorm(20)/5, rnorm(20)/5 )
df <- data.frame(x, variable, value)
d <- ggplot(df, aes(x=x, y=value, group=variable, colour=variable )) +
  geom_line(aes(linetype=variable, colour=variable)) +
  scale_linetype_manual(breaks=lab_strata, values=line_type, labels=LETTERS[1:6]) +
  scale_colour_manual(breaks=lab_strata, values=palette, labels=LETTERS[1:6]) +
  theme(legend.text=element_text(size=5),
        legend.title = element_text(size=6),
        legend.key.height = unit(1/6, 'in'))
d
```

To match what we want, the order of the legends, colours and line types should be like:
```{r,  echo = FALSE, message = FALSE, warning = FALSE}
lab_strata[c(1,3,2,4,6,5)]
```
And the reverse order would be:
```{r,  echo = FALSE, message = FALSE, warning = FALSE}
l <- levels(g$data.survtable$strata)[c(6,2,4,5,1,3)]
l
```
Now we can start with the preliminary plot:

```{r,  echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3, fig.width=5}
g$plot +
  scale_colour_manual(name='', breaks=l, values=palette,
                      labels=lab_strata[c(1,3,2,4,6,5)],
                      guide=guide_legend(nrow = 3)) +
  scale_fill_manual(name='', breaks=l, values=palette,
                    labels=lab_strata[c(1,3,2,4,6,5)],
                    guide=guide_legend(nrow = 3)) +
  scale_linetype_manual(name='', breaks=l, values=line_type,
                        labels=lab_strata[c(1,3,2,4,6,5)],
                        guide=guide_legend(nrow = 3)) +
  theme(legend.text=element_markdown(),
        plot.title = element_markdown())
```

It seems that the only susceptible flies are the _ken^1^_